
---
title: "American Time Use Survey Analysis"
author: "Maria Ramos"
date: "XX/XX/2021"
output: html_document
---
```{r setup, include=FALSE}
## install packages if necessary
list.of.packages <- c("janitor", "ggplot2", "dplyr", "lubridate", "zoo", "forcats", "ggthemes", "knitr", "rmarkdown")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
## load packages
library(janitor)
library(ggplot2)
library(dplyr)
library(lubridate)
library(zoo)
library(forcats)
library(ggthemes)
library(knitr)
library(tidyr)
## set working directory
knitr::opts_knit$set(root.dir = '/Users/zhans01/Documents/cbds/officehr_scratch/capstone/')
```
*Note*: If you try to Knit this document at this time, you *will* get an error because there is code in this document that has to be edited (by you!) before it will be able to successfully knit!
### GitHub Setup
To get started, you'll want to go to GitHub and start a new repository:
- Call this repository `final_project`. 
- Add a short description
- Check the box to "Initialize this repository with a README. 
- Click `Create Repository`
Once the repository has been created, Click on `Clone or download` and copy the "Clone with HTTPS" link provided. You'll use this to clone your repo in RStudio Cloud. Refer to the materials in this course if you're stuck on this part of the project.
### Directory Setup
In your project, create folders with the following structure:
- data/
  - raw_data/
  - tidy_data/
- code/
  - raw_code/
  - final_code/
- figures/
  - exploratory_figures/
  - explanatory_figures/
- products/
  - writing/
Upload the data file into the tidy_data folder and this .Rmd file into the final_code folder.
Once the .Rmd document is in the correct folder, you'll want to **change the author of this document** to your name at the top of the .Rmd document (in the YAML). Save this change before moving to the next step. Refer to the materials in this course if you're stuck on this part of the project.
### The Data
The American Time Use Survey (ATUS) is a time-use survey of Americans, which is sponsored by the Bureau of Labor Statistics (BLS) and conducted by the U.S. Census Bureau. Respondents of the survey are asked to keep a diary for one day carefully recording the amount of time they spend on various activities including working, leisure, childcare, and household activities. The survey has been conducted every year since 2003.
Included in the data are main demographic variables such as respondents' age, sex, race, marital status, and education. The data also includes detailed income and employment information for each respondent. While there are some slight changes to the survey each year, the main questions asked stay the same. You can find the data dictionaries for each year on [https://www.bls.gov/tus/dictionaries.htm](https://www.bls.gov/tus/dictionaries.htm). Specifically, you can find data dictionaries (also called codebooks) at [https://www.bls.gov/tus/atuscpscodebk16.pdf](https://www.bls.gov/tus/atuscpscodebk16.pdf) for the CPS data and at [https://www.bls.gov/tus/atusintcodebk16.pdf](https://www.bls.gov/tus/atusintcodebk16.pdf) for the rest of the variables.
### Accessing the Data
There are multiple ways to access the ATUS data; however, for this project, you'll get the raw data directly from the source. The data for each year can be found at [https://www.bls.gov/tus/#data](https://www.bls.gov/tus/#data). Once there, there is an option of downloading a multi-year file, which includes data for all of the years the survey has been conducted, but **for the purposes of this project, let's just look at the data for 2016**. Under **Data Files**, click on `American Time Use Survey--2016 Microdata files`. 
You will be brought to a new screen. Scroll down to the section **2016 Basic ATUS Data Files**. Under this section, you'll want to **click to download** the following two files: `ATUS 2016 Activity summary file (zip)` and `ATUS-CPS 2016 file (zip)`. 
* `ATUS 2016 Activity summary file (zip)` contains information about the total time each ATUS respondent spent doing each activity listed in the survey. The activity data includes information such as activity codes, activity start and stop times, and locations.
* `ATUS-CPS 2016 file (zip)` contains information about each household member of all individuals selected to participate in the ATUS.
Once they've been downloaded, you'll need to **unzip the files**. Once unzipped, you will see the dataset in a number of different file formats including `.sas`, `.sps`, and `.dat` files. **We'll be working with the .dat files.**
### Importing data
Use the first approach explained above to download and access the ATUS data for 2016. Download the CPS and Activity Summary files in a folder and unzip them and within each folder upload the files ending in .dat to `data/raw_data` filder on RStudio.cloud. To load the data in, **run the code in the `atus-data` code chunk** to create an object called `atus.all`.
```{r atus-data, warning=FALSE}
atus.cps <- read.delim('/cloud/project/data/raw_data/atuscps_2016.dat', sep=",")
atus.sum <- read.delim('/cloud/project/data/raw_data/atussum_2016.dat', sep=",")
atus.all <- atus.sum %>%
  left_join(atus.cps %>% filter(TULINENO==1), by = c("TUCASEID"))
```
### Exploratory Analysis of Child Care Data
By looking at the variables in the data frame atus.all, we see that there are a lot of variables that start with t followed by a 6-digit number. These variables capture the total number of minutes each respondent spent doing each activity. [This](https://www.bls.gov/tus/lexiconwex2016.pdf) lists all the activity codes. Using the information in that file, what column is associated with the activity “playing computer games”? (Your answer should start with t and then a 6-digit number that combines the major category, the 2nd tier, and the 3rd tier of the activity. For instance, if major category in equal to 05, second tier is equal to 03, and third tier is equal to 01, then your answer should be t050301.) 
[*t120307*] 
In the data, the variable t010101 contains the total number of minutes each respondent spent doing activity 010101 which is “sleeping” and the variable t010102 contains the total number of minutes each respondent spent doing activity 010102, “sleeplessness.” Find the variable associated with “Socializing and communicating with others.” How much time, on average, does a person in the sample spend on “Socializing and communicating with others”?
[*38.06481*] 
```{r initial-exploration}
mean(atus.all$t120101)
```
Find the activity code that is associated with “Caring For & Helping HH Children”. This is a level 2 activity so we need to add all the variables that start with t0301. Create a column in the data frame atus.all called CHILDCARE that is the sum of all the columns that start with t0301. Here, you’ll have to change the code chunk creating-childcare-var. `rowwise()` is an operation that comes from `dplyr` to perform row-wise aggregation (e.g. compute sum of x, y, z). Read more about this function [here](https://dplyr.tidyverse.org/articles/rowwise.html).
```{r creating-childcare-var}
atus.all <- atus.all %>%rowwise(TUCASEID) %>%
  mutate(CHILDCARE = sum(c_across(starts_with("t0301"))))
names(atus.all)

#Create a density plot of the variable CHILDCARE using ggplot2:
#```{r childcare-density-plot}
childcare_density_plot <- ggplot(atus.all, aes(CHILDCARE, na.rm = FALSE)) +
    geom_density()
```
From the data dictionary, what is the variable that shows the sex of the respondent? 
[*TESEX*]
From the data dictionary, the variable that represents the gender of the respondents can one of take two values, 1 or 2. Which gender group does 1 represent?
[*MALE*] 
### Regression Analysis
We are going to answer whether women or men spend more time with their children. Start by grouping individuals by their gender and calculate the average time men and women spend with their children. Use the code chunk gender-analysis in the .Rmd file. Note that you should replace FUNCTION in order to calculate the average of the variable CHILDCARE.
Men and women are different in the amount of time they spend with their children. Which group spends more time, men or women?
[type answer here:] 
```{r gender-analysis}
atus.all %>% 
    group_by(TESEX) %>% 
    summarise(avg_parent_childcare=mean(CHILDCARE))
```
Use the table() function to look at the variable TRDPFTPT which shows whether the person works full time or part time. You will notice that the variable also takes the value -1. This is probably due to non-response or other data collection reasons. Replace these values with NA in your data so they don’t affect your analysis. Use the code chunk replacing-na for doing this and add your commands there.
```{r replacing-na1}
## replace -1 in the variable TRDPFTPT with NA.
atus.all$TRDPFTPT[atus.all$TRDPFTPT==-1]<-NA %>%
  ifelse(is.na(atus.all$TRDPFTPT))
```

How many NAs are in the variable now?
[*4119*] 
```
sum(is.na(atus.all$TRDPFTPT))

```
The variable on "Full time labor force" status is PRFTLF which also has NA values that take on the value of -1. Repeat the steps above to replace -1 in the variable PRFTLF:
```{r replacing-na2}
## replace -1 in the variable PRFTLF with NA.
atus.all$PRFTLF[atus.all$PRFTLF==-1]<-NA %>%
  ifelse(is.na(atus.all$PRFTLF))

View(atus.all$PRFTLF)
```

Now, we are going to explore what factors affect time spent with children to answer the following questions:
  
For each of these relationships, present your results in a table or a graph. You can do this by first finding these variables from the data dictionary. Add your code in the code chunk exploratory-analysis in the .Rmd file. *Make sure that in your analysis, you limit your data to those who have at least one child (18 or younger) in the household.* The variable for the number of children (18 or younger) in the household is TRCHILDNUM. Use the data frame atus.all that you previously created and again limit your data to those who have at least one child (18 or younger) in the household. The variable for income is HEFAMINC. Other variables are easy to find from the data dictionary.
1. Do younger parents spend more time with their children?
[*yes younger parents spend more time with their kids*:] 
```{r exploratory-analysis1}
## add your exploratory analysis code here
at_least_one_kid<-atus.all %>%
  select(CHILDCARE, TRYHHCHILD, HEFAMINC, PEMARITL, TRDPFTPT, TEAGE,TRCHILDNUM, TESEX)%>%
    filter(TRCHILDNUM > 0)
 

 
```
2. Do richer people spend more time with their children compared to poorer people
[*income does not affect the time send with their child*] 
```{r exploratory-analysis2}
## add your exploratory analysis code here

  ggplot(at_least_one_kid, aes(x = HEFAMINC, y = CHILDCARE, col = TESEX )) + geom_point(aes(color = factor(TEAGE)), size = 1) + (legend.position = "none") + labs( x = "FAMILY INCOME", y = "CHILDCARE", title = "Do rich people spend more time with \n their children than poor people?")

```


3. Do married couples spend more time with their children compared to single parents?
[*married couples spend more time with kids*] 
```{r exploratory-analysis3}
## add your exploratory analysis code here
ggplot(at_least_one_kid,aes(x=factor(PEMARITL),y=CHILDCARE,fill=factor(PEMARITL))) + geom_bar(stat = "identity")+ labs(x= "marital status", y="CHILDCARE", title = "Do married couples spend more \n time with their children compared to single parents? ")
```
4. Do full-time workers spend more time with their children compared to part-time workers?
[*full time workers spend more time then part time workers*] 
```{r exploratory-analysis4}
## add your exploratory analysis code here
filter(at_least_one_kid,!is.na(TRDPFTPT))%>%
ggplot(aes(x=factor(TRDPFTPT),y=CHILDCARE, color=(TRDPFTPT))) + geom_point() + labs(x="Full/Part Time",y = "CHILDCARE",title="Do full-time workers spend more \n time with their children compared to part-time workers?")

```
jIn the exercise above, we looked at bilateral (two-way) relationships. For instance, we looked at how income and time spent with children are related. You have learned in this course, however, that other confounding variables can be a source of bias in your analysis. For instance, the effect of income on time spent with children can be biased by the number of children a person has. Maybe richer people spend less time because they have fewer children. It’s much better to look at the relationship of all relevant variables associated with time spent with children together. Run a linear regression of marital status, age, sex, number of children (18 or younger), earnings, and full-time versus part-time status. Add your code in the reg-analysis code chunk. *Remember to limit the sample to those who have at least one child (18 or younger) in the household.* Also make sure to to change the values of the variable TRDPFTPT that are -1 to NA.
```{r reg-analysis}
## add your regression analysis code here
at_least_one_kid <- lm(CHILDCARE ~ TEAGE +PEMARITL+HEFAMINC+ TESEX+ TRDPFTPT + TESEX, data = at_least_one_kid)
summary(at_least_one_kid)
```
What is the coefficient on gender now?
[type answer here:] 
In the regression, the coefficient on the variable age means how much time spent with children changes if age increases by 1. Based on your results, what’s the difference in minutes spent with children between two people with 10 years of age difference?
[type answer here:] 
### Exploratory Analysis of Age and Activities
In the next few questions, we are going to see whether time spent on different activities varies by age. However, we will only consider activities at the major category levels. There are 18 major categories in the data including personal care, household activities, caring for & helping household members, etc. Because each activity column in the data is at the 3rd tier, we will need to start by suming columns at the activity major categories levels. Save the resulting data frame as df.wide. Use code chunk activity-cats in the .Rmd file for this part. In addition to `rowwise()`, you may also need to utilize `c_across` and `starts_with` to help save some time from typing! Read more about these two functions [here](https://dplyr.tidyverse.org/reference/c_across.html?q=c%20_%20across) and [here](https://tidyselect.r-lib.org/reference/starts_with.html).
```{r activity-cats}
atus.wide <- atus.all %>%rowwise(TUCASEID)%>%
    mutate(act01 = sum(c_across(starts_with("t01"))),
           act02 = list(rowSums(atus.all[, grep("t02", names(atus.all))])),
           act03 = rowSums(atus.all[,grep("t03", names(atus.all))]),
           act04 = rowSums(atus.all[,grep("t04", names(atus.all))]),
           act05 = rowSums(atus.all[,grep("t05", names(atus.all))]),
           act06 = rowSums(atus.all[,grep("t06", names(atus.all))]),
           act07 = rowSums(atus.all[,grep("t07", names(atus.all))]),
           act08 = rowSums(atus.all[,grep("t08", names(atus.all))]),
           act09 = rowSums(atus.all[,grep("t09", names(atus.all))]),
           act10 = rowSums(atus.all[,grep("t10", names(atus.all))]),
           act11 = rowSums(atus.all[,grep("t11", names(atus.all))]),
           act12 = rowSums(atus.all[,grep("t12", names(atus.all))]),
           act13 = rowSums(atus.all[,grep("t13", names(atus.all))]),
           act14 = rowSums(atus.all[,grep("t14", names(atus.all))]),
           act15 = rowSums(atus.all[,grep("t15", names(atus.all))]),
           act16 = rowSums(atus.all[,grep("t16", names(atus.all))]),
           # act17 = , there is no category 17 in the data
           act18 = rowSums(atus.all[,grep("t18", names(atus.all))])) %>% 
    select(TUCASEID, TEAGE, HEFAMINC, starts_with("act"))

    head(atus.wide)


```
What activity group do people spent, on average, most time on?
[type answer here:] 
Which is the second most time consuming activity for the respondents?
[type answer here:] 
What is the maximum time a person in our sample spends time on activity category 5 (Work & Work-Related Activities)?
[type answer here:] 
Convert the data from wide to long using the package of your choice and save the data frame as df.long. Use the code chunk wide-to-long for this purpose. Make sure your key variable is called ACTIVITY and your value variable is called MINS.
Now, group the data frame you created in the previous step by activity type (ACTIVITY) and age (TEAGE). Calculate the average time for each age group and call it AVGMINS. In ggplot2, plot AVGMINS against TEAGE for each category (multiple panels). Type your code in the code chunk age-activity. Label each panel in your graph with the appropriate activity name.
For which categories does the average time spent vary by age? 
[type answer here:]
```{r wide-to-long}
atus.long <- atus.wide %>% 
  # use code to convert the wide format to long.
```
```{r age-activity}
atus.long %>% 
    group_by(ACTIVITY, TEAGE) %>% 
    summarise(AVGMINS = mean(MINS)) %>% 
    ggplot() +
```
Based on the graph, what is true about the activity category 5 (Work & Work-Related Activities)?
[type your answer here:] 
Based on the graph, what is true about the activity category 12 (Socializing, Relaxing, and Leisure)?
[type your answer here:] 
### **Optional:** Exploratory Analysis of Income and Activities
Finally, in this last step, we are going to create a graph that shows how different income groups spend time doing each activity. The graph is based on Henrik Lindberg’s data visualization posted [here](https://raw.githubusercontent.com/halhen/viz-pub/master/pastime-income/pastime.png). The only difference is that we are only looking at the 18 major activity categories. Use the long data that you created in the previous section and make the graph as close as possible to the graph by Henrik Lindberg. Type your code in the code chunk activity-income.
```{r activity-income}
atus.long %>% 
  group_by(ACTIVITY, HEFAMINC) %>% 
  ## add the rest of the code here
  
```
```{r save-plot}
## save the plot above
```
## Final Steps
Congratulations! You have completed the project. There are a few final notes:
### Add Markdown Text to .Rmd
Before finalizing your project you'll want be sure there are **comments in your code chunks** and **text outside of your code chunks** to explain what you're doing in each code chunk. These explanations are incredibly helpful for someone who doesn't code or someone unfamiliar to your project.
### Knit your R Markdown Document
Last but not least, you'll want to **Knit your .Rmd document into an HTML document**. If you get an error, take a look at what the error says and edit your .Rmd document. Then, try to Knit again! Troubleshooting these error messages will teach you a lot about coding in R.
### A Few Final Checks
A complete project should have:
- Completed code chunks throughout the .Rmd document (your RMarkdown document should Knit without any error)
- Comments in your code chunks
- Answered all questions throughout this exercise.
### Final `push` to GitHub
Now that you've finalized your project, you'll do one final **push to GitHub**. `add`, `commit`, and `push` your work to GitHub. At the end, submit the link to your github repository to us through Slack.
